---
title: "Introduction to data analysis and visualization with R_RStudio. Working directory, importing data, and strings management"
author: "Oscar Montoya"
date: "November 26th, 2019"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_depth: 3
  html_notebook:
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    toc: yes
    toc_depth: '3'
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective
- Learn how to set a working directory
- Import your data into R
- Become familiar with some of R's functions for strings management

At the end of this section you should be able to set a working directory, read comma-separated value files, and manipulate strings.

### Working directory
Setting a working directory (wd) helps you to keep things tidy.

Set the wd using the setwd() command
```{r}
# Always start with a clean environment
rm(list=ls(all=TRUE))

# First, check what is your current directory


# Set working directory


```


Set the wd using RStudio GUI (bottom-left panel)

### Importing comma separated files
```{r}



# if needed can concert Species into factor and write it to iris.dataset

   #Now Species is a factor (categorical data)
```


### Summary statistics

Most common statistical basic functions (e.g. central tendency and dispesion measures) are included in the basic intallation of R (base R)

```{r}

# Read iris dataset and add one NA value




# Get the mean of Sepal.Length
   # The answer is NA
   # Removing NA fixes the issue




#No function for standard error so let's install a package that has it:
#install.packages("plotrix")
library(plotrix)



```

Other functions to inquire about the data
```{r}

     

      # what's going on? The subset data has one column so it's not possible to count the number of columns of a vector





```

#### Logical operators

To find "types" of numbers, %/% and %% are handy

```{r}


# %% returns the remainder


# %/% returns the quotient


#Even numbers


#numbers divisible by 5 **OR** 6


# Numbers that are not divisible by 4

```


Can also use brackets for more complicated selections

```{r}


```


### Problem 
Use logical operators to simultaneously find the odd numbers between -50 and 0 (inclusive) and the even numbers between 0 and 50 (inclusive)

```{r}



```



Problem 1

Triple the vaues for sepal length in iris.dataset. Name the new variable "Sepal.Legth.Triple".
```{r}

# to add a new column:


```


### Character data manipulation:

```{r}
?substring()
?strsplit()
?paste()
?sub()
?gsub()
?grep()
?sapply()
?match()
?unique
```


#### Strings manipulation
```{r}
?as.integer
?c
?"["
?letters



# R is smart (most of the time), for logical operators it ranks character sequences alpha-numerically
```


##### Combine strings
```{r}
?paste

#Use the paste function to combine multiple vectors of character data into a single string


#Can change the seperation using the sep parameter



# can combine multiple objects from a single vector into a single string



#Can combine any one-celled variable


#If you have multi-celled variables, you will create multiple strings

```


##### Separate strings
```{r}

?substring
?nchar
?strsplit
?sub

#substring (aka substr) and strsplit can seperate strings


#substring grabs the characters between two points in the string



#nchar tells you how long the string is



#strsplit seperates the string into fragments
#It makes breaks when it finds the string you want within the larger string



#The string you are breaking on is not included in the output




#strsplit puts its output inside of a list (since the arrays in the list can be of different sizes)
#We get pull the relevant vector/array out of a list using [[NUMBER]]



#We can use sapply if our list has multiple slots and each array is the same size


        # Use any of the characters inside the quoatation marks as string separator


#The ".*X" syntax is a regular expression that is recognizable to perl and other languages. It is matching the X argument after any given string, in this case it is matching on the colon (:) and splitting the string and ignoring everything prior to the colon



# Try a value of 1, and a value of 2 here


# Why does the sapply function work??? Because the command "[" is a command to R to extract information specific to lists and the function 'strsplit' creates an object formatted as a list


```


##### Managing strings

```{r}

?"[["
?substr
?substring



# substring(x, start, stop)



## ^^ Try different values instead of 4, such as 3, 5, 1, etc.



# Other ways to get similar results as previously done



# ^^ Try different values for 1 above



```



##### Matching strings

```{r}
?match
?grep



```


### Saving R sessions and individual objects\
When/why to save entire sessions?
When you have ran long analyis and want to save some time that you open them, given the case that you actually need to keep working the unfinished work

When to save individual objects?
WHen you need to import your object/table or simply save it for later

```{r, eval=FALSE}
# Save the session


# Save individual objects


```


### Saving objects in comma separated value (csv) format
How to save in tab separated, colon separated, etc.?

```{r}


```

### Get information about your session
It's always good practice to be aware and report the packages/versions used in your analysis
```{r}

```
