---
title: "Introduction to data analysis and visualization with R_RStudio. Data visualization with ggplot2"
author: "Oscar Montoya, M.Sc."
date: "December 3rd, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Packages used in this section

```{r, eval=FALSE}
install.packages("reshape2", dependencies = TRUE)
install.packages("cowplot", dependencies = TRUE)
install.packages("extrafontdb", dependencies = TRUE)
install.packages("extrafont", dependencies = TRUE)
install.packages("RColorBrewer", dependencies = TRUE)
install.packages("vegan", dependencies = TRUE)

install.packages("devtools", dependencies = TRUE)
install_github("ggbiplot", "vqv")
```

```{r}
my_packages <- c("reshape2", "datasets", "ggplot2", "cowplot", "extrafont", "extrafontdb", "extrafont", "RColorBrewer", "vegan", "devtools", "ggbiplot")

lapply(my_packages, library, character.only = TRUE)

```


## Plot types

### What is the most suitable type of graphic to show my data?

[Grammar of Graphics ggplot2](http://moderngraphics11.pbworks.com/f/ggplot2-Book09hWickham.pdf)

Take a look at [ggplot2's cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) to help you find an aswer to that question.


#### What makes good graphics?

[A layered grammar of graphics](https://byrneslab.net/classes/biol607/readings/wickham_layered-grammar.pdf), by [Hadley Whickham](http://hadley.nz/)

Recommended read: [PLots to avoid](http://genomicsclass.github.io/book/pages/plots_to_avoid.html).

[grammar of graphics](http://moderngraphics11.pbworks.com/f/ggplot2-Book09hWickham.pdf)


#### Plots to avoid
For each bad plot there's a better option. The table below summarises the most common "bad plots" (plots to avoid) and their corresponding "better" alternative. 

Plot to avoid               |   Alternative 
--------------------------- | --------------
Pie chart                   |    Bar plot 
Donut plots                 |    Bar plot 
Bar plots as data summaries |    Box plot
Regressions without the scatter | Regression with scatter
3D plots for 2D data | 2D plots
-------------------------------------------
Table: Plots to avoid and recommended alternatives. \label{tab:plots_to_avoid}


## Introduction to ggplot2 

### ggplot2 and the grammar of graphics philosophy behind it

What Is Grammar Of Graphics?

"The basic idea: independently specify plot building blocks and combine them to create just about any kind of graphical display you want". Those blocks are:

* **data**: must be a **data frame** for ggplot2 to work
* **aesthetic mapping (aes)**: describes the relationship (x and y, colors, fill) but doesn't supply data.
* **geometric object (geom)**: Shapes (e.g. circles and diamonds) that represent observations
* **statistical transformations (stats)**: Some geoms are indeed statistics, e.g. geom_histogram().
* **scales**: Linear scaling of x and y axes. Scales discrete and continuous.
* **coordinate system**: e.g. geom_smooth(), stat_summary()
* **position adjustments**:
* **faceting**: make multipanel plots of data subsets. Helpful to "declutter" busy graphics

"Every geom has a default statistic, every statistic has a default geom (can be overridden)" (http://ggplot2.org/resources/2007-vanderbilt.pdf).

Based on these concepts, ggplot2 makes fairly easy to produce publication quality plots with only a few lines of code.



# Plots examples
## Data: Iris dataset

R has many available datasets, most of them part of the "datasets" package which is part of base R packages.


Load the "datasets" package first, the call the dataset using the "data" function
```{r}
library(datasets) 
data("iris")
```


The first thing to do (always!) is to take a look at the strcuture of your data: Check for potential missing data (NAs, NaNs), typr of object (data frame, matrix), data types (factor, numeric, character), etc.
```{r}

head(iris) # first six rows
str(iris) # structure of each variable (column) in the dataset
class(iris) # Object class
summary(iris) # summary statistics
which(is.na(iris))

test <- iris
test[2,2] <- NA # Turn observation (row) 2 from variable (column) 2 into NA
which(is.na(test))
```



## Histograms

Histograms in base R
```{r}


```

Histograms in ggplot
```{r}
library(ggplot2)


```


## Boxplots

Base R

```{r}

```

ggplot

ggplot automatucally optimizes scales for imporved aesthetics, ultimatelly a graphic that is easier to interpret.
```{r}

```


## Scatter plot

```{r}


```

ggplot
```{r}

```


## A closer look:

Histogram:
* Geom/Stat: Histogram
* Scale: linear
* Coordinate system: cartesian

Bar plot:
* Geom: bar
* Stat: bin
* Scale: linear
* Coordinate system: cartesian

Scatter plot:
* Geom: point
* Stat: identity
* Scale: linear
* Coordinate system: cartesian

```{r, fig.height = 6, fig.width = 12, fig.align = "left"}
library(cowplot)

```


## Facet grids

Facets are handy to avoid busy, hard to interpret plots.


```{r}

# Horizontal facets



# Horizontal facets (default) 



#Vertical facets


```


## Stats

### Stats smooth
```{r}

# Notice the colors argument (inside aes()!)



```


A common mistake is to put the color argument outside aes()
```{r}



```



## Scales

Scales are used to modify axes and colors. 

For example:

* `facet(aes(scale = "free"))`

```{r}
     # notice this is facet_grid, no facet_wrap. facet_grid() makes plots for all possible result/ecombinations in the ploted variables, including empry plots. facet_wrap() only makes plots that contain data.
```

* `scale_y_continuous()` Set name, breaks, labels, limits of y-axis


```{r}




```


* `scale_x_log10()` log transform the x-axis

```{r}




```


* `scale_colour_manual()` Specify colors for geoms

```{r}



```


* Change the order of the elements in the *legend* (deafult is alphabetical)

```{r}



```


## Other types of geoms available in ggplot2:
```{r}


```


## Heatmaps
```{r}
library(ggplot2)




```



## Other common types of plots

```{r}

# horizontal lines




      # regression lines

# violin plots (similar to box plots but also show the probability density)




# Vertical lines




# Add line segments instead of a full line (use with package "grid" to form arrows)



          # whatever the curve and segment mean to you. Notice they're included in the legend.


# Add lines



# Add elipses (stat)



```


## Build a graphic from scratch


### Line/scatter plot


```{r, fig.height = 9, fig.width = 12, fig.align = "left"}
library(extrafont)
fonts()
#loadfonts() # fonts need to be loaded to work with ggsave()


#library(extrafontdb)
#detach("package:extrafont", unload=TRUE)
#detach("package:extrafontdb", unload=TRUE)






```

## Save your plot with ggsave
```{r}

ggsave(filename = "my_plot.pdf", device = "pdf",
       plot=my_plot, height=20, width=24, units="cm")

warnings()
```


### Challenge

Make your own scatter plot using the "economics" dataset. Use pop as your reponse varialble. Apply the themes used for the bar graph from above.

```{r, eval=FALSE, include=FALSE, echo=FALSE}
head(economics);str(economics)




```



## Ordination plots
### Multidimiensional scaling (MDS)


```{r}
library(vegan)

  


 # herb.comm = community od interest, trace=non-verbose output, autotransform = prevents vegan undesired data transformations


?metaMDS
# Available distances:
# "manhattan", "euclidean", "canberra", "bray", "kulczynski", "jaccard", "gower", "altGower", "morisita", "horn", "mountford", "raup" , "binomial", "chao", "cao" or "mahalanobis".

# Access the vegan object with $ symbol and extract the actual data (called points)




# Make a data frame with the generated data and 



# Import data frame into ggplot2
library(ggplot2)




```


### Principal component analysis

```{r}
library(devtools)
#install_github("ggbiplot", "vqv")


# Prepare the data
head(iris)

# Log transformation



library(ggbiplot)


```


PCA: uses inly Euclidian distances (no missing values allowed)
NMDS: robust to missing/null values

Remember: these are only ordination methods, so you still need to perform tests to find about those singinificant differences beteween your samples, treatments, etc.





```{r}
sessionInfo()
citation("ggplot2");citation("cowplot"); citation("extrafont")

```


Recommended addition packages:
RColorbrewer (includes color-blind friendly palettes)
Cowplot (linear and non-linear time series)
scales (easy manipulatiion of scales)

## Sources: 
* https://rpubs.com/mccannecology/53464
* http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html
* https://learnr.wordpress.com/2010/01/26/ggplot2-quick-heatmap-plotting/
* https://www.r-bloggers.com/how-to-create-a-fast-and-easy-heatmap-with-ggplot2/