---
title: "6. Flow control"
author: "Oscar Montoya"
date: "October 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## for loops

R doesn't care if you write something 1000 times or have it copy it 1000 times
But the second is a lot easier for you

```{r}

```

for loops have two stages
1. Set a variable to equal the next value in a sequence
2. Run a set of code with that variable at that value
3. Repeat

```{r}

```

R either takes the rest of the line, the next line, or everything within { } as the looped code

```{r}




```

We can use the variable simply as a tracker to repeat code a given number of times
For example, we can replicate the rep() function

```{r}


```


## Cycle through values

The most useful thing to do with a for loop is to cycle through values

```{r}




```


#### Example
Use t-test to look for methane production diffrences between the salinity levels of every 10th day (starting at day 2), excluding the "BRACKISH" salinity.

```{r}
#subset dataset for every 10th day and exclude brackish

```



## if statements

The big advantage of programming is you can have R make decisions
Allows code to be flexible, dont have to rewrite everything if you change the data
R makes binary decisions: "if data meets a condition, do this thing or else do this other thing"
if(conditional argument that leads to one F or T outcome)

```{r}


```

R either takes the rest of the line, the next line, or everything within { } as the conditional action

```{r}


```

if() is useful if we have R make the decision for us

```{r}


```

We can use the else statement for efficiency if we want an alternative thing to be done only when the first thing is not done else goes after the if() task and runs using the same rules as an if() task

```{r}


```

For simple tasks, ifelse() puts it all in one function
Need to use <- instead of =

```{r}



```


## While and repeat loops

```{r}

```


## Next and break

```{r}

```


## Debugging loops

```{r}
#Sy.sleep can help you debug your for loops by slowing it down (if combined with a print or plot function)



#print can help you figure out where you loop is breaking down


      #you can alway print the variables in your for loop to see the value they reached when it stopped working or look at it's value in the environment

      #you can also set the variable you are looping with to different value to see what happens when you run lines of codes within your loop

#it can really helps to manually run through the loop line by line
```

